---
title: "Normal Distribution"
author: Abdullah Al Mahmud
date: \today
output: 
  revealjs::revealjs_presentation:
    theme: sky
    reveal_plugins: ["chalkboard", "zoom", "menu"]
    self_contained: false
    lib_dir: libs
    css: css/styles.css
    includes:
      in_header: css/header.html
---

# Most Important

It's the 

- most important distribution
- most common distribution
- most widely-used distribution

## Shape

```{r normal-density, echo=FALSE, fig.width=7, fig.height=5, fig.cap=""}
set.seed(123)         # for reproducibility
n <- 1000             # number of samples
mu <- 75               # mean
sigma <- 10            # standard deviation

# simulate data
x <- rnorm(n, mean = mu, sd = sigma)

# plot histogram scaled to density
hist(x,
     breaks = 30,
     probability = TRUE,
     col = "lightblue",
     border = "white",
     main = "Histogram with Normal Density Curve",
     xlab = "Value",
     ylim = c(0, max(density(x)$y, dnorm(mu, mu, sigma)) * 1.1))

# add theoretical normal density
curve(dnorm(x, mean = mu, sd = sigma),
      from = min(x) - 1,
      to = max(x) + 1,
      col = "red",
      lwd = 2,
      add = TRUE)

# add kernel density of the sample
lines(density(x), col = "darkgreen", lwd = 2, lty = 2)

# legend
legend("topright",
       legend = c("Theoretical normal", "Sample density"),
       col = c("red", "darkgreen"),
       lwd = 2, lty = c(1, 2))
``` 

## PDF

The normal PDF is given by

$$
f(x) = \frac{1}{\sigma\sqrt{2\pi}} e^{-\frac{1}{2}\left(\frac{x-\mu}{\sigma}\right)^2}; -\infty <x<\infty
$$

> - First discovered by Abraham De Moivre

## Binomial to Normal

> - Number of trial (n) is large ($n \rightarrow \infty$)
> - $P(S) \approx P(\bar S$), where $S =$ Success

[Check out here](https://portal.statmania.info/ds/analysis/viz-dist.html)

## Poisson to Normal

$m \rightarrow \infty$

$$
\lim_{m \to \infty} \text{Pois}(m) \approx \mathcal{N}(m, m)
$$
[Check out here](https://portal.statmania.info/ds/analysis/viz-dist.html)

## Empirical Rule

68‑95‑99.7 Rule

```{r empirical-rule-base, echo=FALSE, fig.width=6, fig.height=5, fig.cap=""}
# 1. Numeric verification
p1 <- pnorm(1) - pnorm(-1)
p2 <- pnorm(2) - pnorm(-2)
p3 <- pnorm(3) - pnorm(-3)


# 2. Visual plot
x <- seq(-4, 4, length = 400)
y <- dnorm(x)
plot(x, y, type="l", lwd=2, col="blue",
     ylab="Density", xlab="Z",
     main="Standard Normal with ±1,2,3σ Shading")

# Function to shade region between -n and +n
shade_region <- function(n, col) {
  ix <- x >= -n & x <= n
  polygon(c(-n, x[ix], n), c(0, y[ix], 0),
          col = col, border = NA)
}

# Shade ±1, ±2, ±3
shade_region(3, rgb(0,0,1,0.1))
shade_region(2, rgb(0,0,1,0.2))
shade_region(1, rgb(0,0,1,0.3))

# Add legend
legend("topright",
       legend = c("±1σ ~68%", "±2σ ~95%", "±3σ ~99.7%"),
       fill = c(rgb(0,0,1,0.3), rgb(0,0,1,0.2), rgb(0,0,1,0.1)),
       border = NA)
```

## Empirical Rule Area

> - $P(\mu - \sigma \le \mu + \sigma) = 0.6826$
> - $P(\mu - 2 \sigma \le \mu + 2 \sigma) = 0.9544$
> - $P(\mu - 3 \sigma \le \mu + 3 \sigma) = 0.9973$

## Properties

> - Bell-shaped 
> - Symmetrical about mean
> - $\beta_1 = 0, \beta_2 = 3$
> - $\mu = Me = Mo$
> - Area under whole curve is 1
> - The curve on either side of mean extends up to infinity.
> -  All odd central moments are zero

## More Properties

> - Empirical rule
> - Linear combination of Normal $\rightarrow$ normal
> -  $MD(\bar X) \frac 45\sigma$
> - $QD = \frac 23\sigma$
> - $M_X(t) = \mathbb{E}[e^{tX}] = \exp\left(\mu t + \frac{1}{2} \sigma^2 t^2\right) -\infty \le t \le \infty$

## Odd Central Moments

$\mu_1 = \mu_3 = \cdots = \mu_{2n+1} = 0$

Mathematically Why?
> - 

## Application or Normal Distribution

> - Quality Control in industry
> - Most distributions tend to be normal under some assumptions
> - Sometimes transformation makes normal
> - Used in hypothesis testing (many samples are normally distributed)

## Standard Normal

If $X \sim N(\mu, \sigma ^2), z = \frac{x-\mu}{\sigma} \sim N(0,1)$

> - $\displaystyle f(z) = \frac{1}{\sqrt{2 \pi}}e^{\frac{-z^2}{2}}$ 
> - $E(Z) = 0$
> - $V(Z) = SD(Z) = 1$

## Standard Normal Mean

Let $X \sim N(\mu, \sigma ^2)$

> - Now let $\displaystyle Z = \frac{X-\mu}{\sigma}$
> - $E(Z) = E(\frac{X-Z}{\mu})$
> - = $\frac{1}{\sigma} E(X-\mu)$
> - = $\frac{1}{\sigma} [E(X) - E(\mu)]$
> - = $\frac{1}{\sigma} (\mu - \mu)$
> - 0

## Standard Normal Variance

$V(Z) = V(\frac{X-Z}{\mu})$

> - = $\frac{}{}$ 

# Normal Problems

## Find Skewness

Find the coefficient of skewness of normal distribution with variance 4.

<div class="container">
<div class="col">
> - $\mu_3 = 0$
> - $\sigma^2 =\mu_2 = 4$
</div>
<div class="col">
> - $\gamma_1 = \sqrt{\beta_1} = \frac{\mu_3}{\sqrt{\mu_2^3}}$
> - $\gamma_1 = \frac{0}{\sqrt{4^3}} = 0$
> -  So no skew (symmetric)
</div>
</div>

## Probablity of Z [-1, 1]


Find $P(-1 \le z \le 1)$

<div class="container">
<div class="col">
<small>

> - $P(-1 \le z \le 1)$
> - $P(-1 \le z \le 0) + P(0 \le z \le 1)$
> - $0.3413 + 0.3413$ 
> - 0.6826

</small>
</div>
<div class="col">

```{r normal_plot_base_R, echo=FALSE, fig.width=7, fig.height=5}
# Define the range of Z-values
z <- seq(-4, 4, length.out = 500)

# Calculate the probability density for each Z-value
density_values <- dnorm(z, mean = 0, sd = 1)

# Plot the standard normal curve
plot(z, density_values, type = "l",
     main = "Standard Normal Distribution: P(-1 <= Z <= 1)",
     xlab = "Z-score",
     ylab = "Probability Density",
     col = "steelblue",
     lwd = 2,
     ylim = c(0, max(density_values) * 1.1), # Extend y-axis slightly for labels
     axes = FALSE) # Suppress default axes to draw custom ones

# Add custom x-axis
axis(side = 1, at = c(-3, -2, -1, 0, 1, 2, 3), labels = c("-3", "-2", "-1", "0", "1", "2", "3"))
# Add custom y-axis (optional, often hidden for density plots)
# axis(side = 2)

# Shade the area between -1 and 1
# Find the indices corresponding to z = -1 and z = 1
idx <- which(z >= -1 & z <= 1)
# Create polygon coordinates
polygon(x = c(z[idx[1]], z[idx], z[idx[length(idx)]]),
        y = c(0, density_values[idx], 0),
        col = "lightblue",
        border = NA) # No border for the shaded area

# Add vertical lines at -1, 0, and 1
abline(v = 0, lty = 2, col = "gray50") # Mean (0)
abline(v = -1, lty = 1, col = "red", lwd = 1.5) # -1 standard deviation
abline(v = 1, lty = 1, col = "red", lwd = 1.5) # 1 standard deviation

# Add text labels for -1, 0, and 1
text(x = 0, y = -0.02, labels = "0", col = "black", pos = 1, cex = 1.2)
text(x = -1, y = -0.02, labels = expression(bold("-1")), col = "red", pos = 1, cex = 1.2)
text(x = 1, y = -0.02, labels = expression(bold("1")), col = "red", pos = 1, cex = 1.2)

# Indicate -infinity and +infinity conceptually
# This is typically implied by the curve extending to the edges of the plot
# You could add arrows, but it often clutters the plot more than it helps for this specific question.
# For example:
# arrows(x0 = -3.5, y0 = 0.01, x1 = -Inf, y1 = 0.01, length = 0.1, col = "darkgreen")
# arrows(x0 = 3.5, y0 = 0.01, x1 = Inf, y1 = 0.01, length = 0.1, col = "darkgreen")
# text(x = -3.5, y = 0.02, labels = expression(-infinity), pos = 4, col = "darkgreen")
# text(x = 3.5, y = 0.02, labels = expression(+infinity), pos = 2, col = "darkgreen")
```
</div>
</div>


## Method 2 

> - $P(-1 \le z \le 1) = P(z\le1) - P(z \le -1)$
> - $P(z \le -1) = 1- P(z \le 1)$ (since symmetric)
> - $P(-1 \le z \le 1) =  P(z\le1) - [1- P(z \le 1)]$
> - $2 P(z \le 1) - 1$
> - $2 \times 8413-1 = 0.6826$

## Find MD of N(20, 25)

<div class="container">
<div class="col">
> - $\mu = 20 , \sigma^2 = 25$
> - $\sigma = 5$
</div>
<div class="col">
Mean Deviation about Mean

> - $MD(\bar X) = \frac45 \sigma$
</div>
</div>

## Find Varinace 

Find the variance of normal distribution when $\mu_4 = 2$

> - Kurtosis, $\beta_2 = \frac{\mu_4}{\mu_2}$
> - Mesokortic $\rightarrow \beta_2 = 3$
> - $3 = \frac{2}{\mu_2^2}$
> - $\mu_2^2 = 0.67$
> - $\sigma^2 = ?$
> - $\sqrt{0.67}$

## P(X > 20)

$\mu = 15, \sigma = 4, P(X > 20)= ?$

<div class="container">
<div class="col">
> - $P(X > 20)$
> - $P(\frac{x-\mu}{\sigma} > \frac{20-15}{4})$
> - $P(z>1.25)$
> - $P(0\le z \le \infty) - P(0\le z \le 1.25)$
> - $0.5 - 0.3944 = 0.1056$
</div>
<div class="col">

```{r normal_plot_base_R_right_tail_1.25, echo=FALSE, fig.width=7, fig.height=5}
# Define the range of Z-values
z <- seq(-4, 4, length.out = 500)

# Calculate the probability density for each Z-value
density_values <- dnorm(z, mean = 0, sd = 1)

# Define the specific Z-score for the boundary
boundary_z <- 1.25

# Plot the standard normal curve
plot(z, density_values, type = "l",
     main = expression(bold(paste("Standard Normal Distribution: P(Z > 1.25)"))),
     xlab = "Z-score",
     ylab = "Probability Density",
     col = "steelblue",
     lwd = 2,
     ylim = c(0, max(density_values) * 1.1), # Extend y-axis slightly for labels
     axes = FALSE) # Suppress default axes to draw custom ones

# Add custom x-axis
axis(side = 1, at = c(-3, -2, -1, 0, 1, boundary_z, 2, 3),
     labels = c("-3", "-2", "-1", "0", "1", expression(bold("1.25")), "2", "3"))
# Add custom y-axis (optional, often hidden for density plots)
# axis(side = 2)

# Shade the area to the right of boundary_z
# Find the indices corresponding to z >= boundary_z
idx <- which(z >= boundary_z)
# Create polygon coordinates: start at boundary_z (z[idx[1]]), go along curve (z[idx], density_values[idx]), then down to 0 at the last point
polygon(x = c(z[idx[1]], z[idx], tail(z, 1)),
        y = c(0, density_values[idx], 0),
        col = "lightcoral",
        border = NA) # No border for the shaded area

# Add vertical lines at 0 and boundary_z
abline(v = 0, lty = 2, col = "gray50") # Mean (0)
abline(v = boundary_z, lty = 1, col = "red", lwd = 1.5) # Z = 1.25

# Add text labels for 0 and boundary_z
text(x = 0, y = -0.02, labels = "0", col = "black", pos = 1, cex = 1.2)
# The label for boundary_z is already included in the axis labels for bolding

# Indicate +infinity conceptually with an arrow and text
arrows(x0 = 3.2, y0 = dnorm(3.2), x1 = 3.8, y1 = dnorm(3.8), length = 0.1, col = "darkgreen", lwd = 1)
text(x = 3.8, y = dnorm(3.8) + 0.02, labels = expression(+infinity), pos = 2, col = "darkgreen", cex = 1)
```

</div>
</div>

## Exam Score

Scores on the GMAT are roughly normally distributed with a mean of 527 and a standard deviation of 112. What is the probability of an individual scoring above 500 on the GMAT?[[source](https://www.stat.colostate.edu/inmem/gumina/st201/recitation8/downloads/Normal%20Probabilites%20Practice.pdf?ref=docs.statmania.info)]

> - $\mu = 527, \sigma = 112$ 
> - $P(X > 500)$

## Gestation

<div style="font-size:30px;">
The length of human pregnancies from conception to birth approximates a normal distribution with a mean of 266 days and a standard deviation of 16 days. 

> - What proportion of all pregnancies will last between 240 and 270 days (roughly between 8 and 9 months)?
> - What length of time marks the shortest 70% of all pregnancies? 
</div>

## Forest burn

<div style="font-size:30px;">

The average number of acres burned by forest and range fires in a large New Mexico county is 4,300 acres per year, with a standard deviation of 750 acres. The distribution of the number of acres burned is normal. 

> - What is the probability that between 2,500 and 4,200 acres will be burned in any given year?

</div>